Short-time expansion of Heisenberg operators in open collective quantum spin systems

### 2019-09-12 soc_squeezing/collective_paper/collective_spin.tex
minor additions/additions to text concerning Heisenberg operators

\begin{document}

@@

\section{Introduction}

@@

\section{Theory}
\label{sec:theory}

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider a system of $N$ distinct
spin-1/2 particles.  Defining individual spin-1/2 operators
$\hat s_{\a=\x,\y,\z}\equiv\hat\sigma_\a/2$ and
$\hat s_\pm\equiv\hat s_\x\pm i\hat s_\y=\hat\sigma_\pm$ with Pauli
operators $\hat\sigma_\a$, we denote an operator that acts with
$\hat s_\a$ on the spin indexed by $j$ and trivially (i.e.~with the
identity $\hat\1$) on all other spins by $\hat s_\a^{(j)}$.  We then
define the collective spin operators
$\hat S_\a\equiv\sum_{j=1}^N\hat s_\a^{(j)}$ for
$\a\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\hat\S_{\v m}}$
as a basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\hat\S_{\v m}\equiv \hat S_+^{m_+} \hat S_\z^{m_\z} \hat S_-^{m_-}$,
we can expand any collective spin operator $\hat\O$ in the form
\begin{align}
  \hat\O = \sum_{\v m} \O_{\v m} \hat\S_{\v m}
\end{align}
with scalar coefficients $\O_{\v m}\in\C$.  If $\hat\O$ is
self-adjoint, for example, then $\O_{\v m}^*=\O_{\v m^*}$ with
$\v m^*\equiv\p{m_-,m_\z,m_+}$.
%
% The corresponding Heisenberg operator is then
% $\hat\O\p{t}=\sum_{\v m}\O_{\v m}\p{t}\hat\S_{\v m}$, with
% time-dependent coefficients $\O_{\v n}\p{t}$ for time-independent
% (Schr\"odinger) operators $\hat\S_{\v m}$.  The time evolution of
% Heisenberg operators is determined by
%
The corresponding Heisenberg operator is then
$\hat\O\p{t}=\sum_{\v m}\O_{\v m}\p{t}\hat\S_{\v m}+\hat\E_\O\p{t}$,
with time-dependent coefficients $\O_{\v n}\p{t}$ for time-independent
(Schr\"odinger) operators $\hat\S_{\v m}$, and mean-zero ``noise''
operators $\hat\E_\O\p{t}$ that result from interactions between the
spin system and its environment, initially $\hat\E_\O\p{0}=0$.  These
noise operators will essentially play no role in the present work, but
are necessary to include for a consistent formalism of Heisenberg
operators in an open quantum system; see Appendix \ref{sec:noise} for
further discussion.  The expectation values of Heisenberg operators
evolve according to
%
\begin{align}
  \f{d}{dt} \bk{\hat\O\p{t}}
  = \bk{\check T \hat\O\p{t}}
  = \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T_{\v m\v n} \O_{\v n}\p{t}
  \label{eq:EOM}
\end{align}
with a Heisenberg-picture time derivative operator $\check T=d/dt$
whose matrix elements $T_{\v m\v n}\in\C$ are defined by
\begin{align}
  \check T \hat\S_{\v n} \equiv i \sp{\hat H, \hat\S_{\v n}}_-
  + \sum_\J \gamma_\J \check\D\p{\J} \hat\S_{\v n}
  = \sum_{\v m} \hat\S_{\v m} T_{\v m\v n},
  \label{eq:time_deriv}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\hat H$ is the collective spin
Hamiltonian; $\J$ is a set of jump operators with a corresponding
decoherence rate $\gamma_\J$; and $\check\D$ is a Heisenberg-picture
dissipator, or Lindblad superoperator, defined by
\begin{align}
  \check\D\p{\J} \hat \O
  \equiv \sum_{\hat J\in\J} \p{\hat J^\dag \hat \O \hat J
    - \f12\sp{\hat J^\dag \hat J, \hat \O}_+}.
\end{align}
Decoherence via uncorrelated decay of individual spins, for example,
would be described by the set of jump operators
$\J_-\equiv\set{\hat s_-^{(j)}:j=1,2,\cdots,N}$.  The commutator in
Eq.~\eqref{eq:time_deriv} can be computed by expanding the product
$\hat\S_{\v\ell}\hat\S_{\v m} = \sum_{\v n} f_{\v\ell\v m\v
  n}\hat\S_{\v n}$ with structure constants
$f_{\v\ell\v m\v n}\in\mathbb{R}$ that we work out in Appendices
\ref{sec:identities}--\ref{sec:general_product}, and the effects of
decoherence from jump operators (i.e.~elements of $\J$) of the form
$\hat g^{(j)} = \sum_\a g_\a \hat s_\a^{(j)}$ and
$\hat G = \sum_\a G_\a \hat S_\a$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be some of the main technical
contributions of this work, with potential applications beyond the
short-time simulation method presented here.  These ingredients are
sufficient to compute matrix elements $T_{\v m\v n}$ of the time
derivative operator $\check T$ in Eq.~\eqref{eq:time_deriv} in most
cases of practical interest.

@@

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \bk{\hat\O\p{t}}
  = \bk{e^{t\check T} \hat \O\p{0}}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator $\check T^k$ are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \bk{\hat\O\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
%
% We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
% expansion of Heisenberg operators.  Unlike the truncation in
% Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
% $T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in Eq.~\eqref{eq:TST} tell us
% which operators are relevant for computing the expectation value
% $\bk{\hat\O\p{t}}$ to a fixed order $M$.  Note that using the
% relation $\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian
% conjugation of Eq.~\eqref{eq:EOM} also implies that
% $T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can cut both the number of
% initial-time expectation values $\bk{\hat\S_{\v m}}_{t=0}$ and the
% number of matrix elements $T_{\v m\v n}$ that we may need to
% explicitly compute roughly in half.
%
We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Note that when computing an
expectation value $\bk{\hat\O\p{t}}$, the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, cuts both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}$ and the number of
matrix elements $T_{\v m\v n}$ that we may need to explicitly compute
roughly in half.

Unlike the weight-based truncation in
Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
$T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in Eq.~\eqref{eq:TST} tell us
which operators $\hat\S_{\v m}$ are relevant for computing the
expectation value $\bk{\hat\O\p{t}}$ to a fixed order $M$.  The TST
expansion thereby avoids the introduction of a weight measure $w$ that
chooses which operators to keep track of, and trades the cost of
solving a system of differential equations for the cost of computing
expectation values $\bk{\hat\S_{\v m}}$ and matrix elements
$T_{\v m\v n}^k$.  In all cases considered in this work, we find that
the TST expansion is both faster to evaluate and provides accurate
correlators $\bk{\hat\O\p{t}}$ until later times $t$ than the
weight-based expansion in \eqref{eq:weight_truncation} with weight
measure $w\p{\v m}=\abs{\v m}$ and cutoff $W\approx M$.  We therefore
restrict the remainder of our discussions to the TST expansion in
Eq.~\eqref{eq:TST}, and provide a pedagogical tutorial for computing
correlators using the TST expansion in Appendix \ref{sec:tutorial}.
%

%
% In principle, the factorial suppression of terms at higher orders of
% the expansion in Eq.~\eqref{eq:time_series} implies for any given
% time $t$, there exists a truncation order $M_t$ for which the
% truncation error in Eq.~\eqref{eq:TST} is negligibly small.  In
% practice, only a maximal truncation order $M_{\t{max}}$ is
% accessible with limited computational resources, such that the TST
% expansion in Eq.~\eqref{eq:TST} only allows us to compute the
% expectation value $\bk{\hat\O\p{t}}$ to a maximal time
% $t_{M_{\t{max}}}$.  As we will see in the following section,
% $M_{\t{max}}=35$ will suffice for the computation of collective spin
% correlators up to times that are relevant e.g.~for spin squeezing
% protocols.  Appendix \ref{sec:tutorial} provides a pedagogical
% tutorial for computing correlators using the TST expansion.
%
Three primary considerations limit the maximum time $t$ to which we
can accurately compute a correlator $\bk{\hat S_{\v n}\p{t}}$ using
the TST expansion.  First, maintaining accuracy at larger times $t$
requires going to higher orders $M$ in the TST expansion.  An
order-$M$ TST expansion of the correlator $\bk{\hat S_{\v n}\p{t}}$
can involve a significant fraction of operators $\hat S_{\v m}$ with
weight $\abs{\v m}\lesssim M$, which implies the need to compute
$O\p{M^3}$ initial-time expectation values $\bk{\hat S_{\v m}}$ and
$O\p{M^4}$ matrix elements $T^k_{\v m\v n}$.  In practice, with a
straightforward implementation of the TST expansion we find that these
requirements generally restrict $M\lesssim 50$ -- $70$ with $8$ --
$50$ gigabytes of random access memory (RAM).  Second, individual
terms at high orders of the TST expansion in Eq.~\eqref{eq:TST} can
grow excessively large, greatly amplifying any numerical errors and
thereby spoiling cancellations that are necessary to arrive at a
physical value of a correlator, i.e.~with
$\abs{\bk{\hat S_{\v n}\p{t}}}\lesssim S^{\abs{\v n}}$.  Finally, the
TST expansion is essentially perturbative in the time $t$, which
implies that the validity of this a formal expansion eventually breaks
down.  Precisely characterizing the implications of these last two
considerations for the TST expansion requires additional analysis that
we defer to future work.  As we show from benchmarks of the TST
expansion in Section \ref{sec:squeezing}, however, a detailed
understanding of these limitations is not necessary to diagnose the
breakdown time $t_{\t{break}}^{(M)}$ beyond which the TST expansion
yields inaccurate results.  Empirically, we find that going beyond
order $M\approx35$ yields no significant gains in all cases considered
in this work.
%

\section{Spin squeezing, benchmarking, and breakdown}
\label{sec:squeezing}

@@

Figure \ref{fig:benchmarking} compares the squeezing parameter $\xi^2$
for $N=10^4$ spins initially in the state $\ket{\X}$ evolved under the
Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}, as computed via
both benchmarking simulations and the TST expansion in
Eq.~\eqref{eq:TST} with $M=35$.  Squeezing is shown for both unitary
dynamics (Figure \ref{fig:squeezing_unitary}), as well as non-unitary
dynamics in the presence of spontaneous decay, excitation, and
dephasing of individual spins at rates $\chi$ (Figure
\ref{fig:squeezing_incoherent}), respectively described by the sets of
jump operators $\J_\a\equiv\set{\hat s_\a^{(j)}}$ with corresponding
decoherence rates $\gamma_\a=\chi$ for $\a\in\set{-,+,\z}$.  The
results shown in Figure \ref{fig:benchmarking} were computed in a
rotated basis with
$\p{\hat s_\z,\hat s_\x}\to\p{\hat s_\x,-\hat s_\z}$ and
$\ket\X\to\ket{-\Z}\equiv\ket\dn^{\otimes N}$, as well as appropriate
transformations of the Hamiltonian and jump operators.
%
% The only effect of this rotation on the results presented in Figure
% \ref{fig:benchmarking} is to prolong the time for which the TST
% expansion agrees with the benchmarking simulations.  The reason for
% different results in rotated basis has to do with the breakdown of
% the TST expansion, which we discuss further below.
%
The only effects of this rotation on the results presented in Figure
\ref{fig:benchmarking} are to
\begin{enumerate*}
\item reduce the time it takes to compute correlators
  $\bk{\hat\O\p{t}}$ with the TST expansion, and
\item prolong the time for which the TST expansion of TNT results
  agree with benchmarking simulations.
\end{enumerate*}
The speedup in a different basis occurs because for the initial state
$\ket{-\Z}$, all initial-time correlators $\bk{\hat S_{\v m}}$ are
zero unless $m_+=m_-=0$, and all non-zero correlators take $O\p{1}$
(i.e.~constant in $N$) time to compute, rather than $O\p{N}$ time (see
Appendix \ref{sec:initial_conditions}).  In total, the use of a
rotated basis reduces the computation time of initial-time correlators
from $O\p{M^3N}$ to $O\p{M}$.  The reason for prolonged agreement of
TNT results in a rotated basis is not entirely understood, and
provides a clue into the precise mechanism by which the TST expansion
breaks down (discussed below).  We defer a detailed study of this
breakdown to future work.
%

The main lesson from Figure \ref{fig:benchmarking} is that the TST
expansion yields essentially exact results right up until a sudden and
drastic departure that can be diagnosed by inspection.  The breakdown
of the TST expansion in Figure \ref{fig:benchmarking} is marked by an
unphysical squeezing parameter $\xi^2<0$, which occurs due to
individual correlators taking unphysical values with
$\abs{\bk{\hat\S_{\v n}\p{t}}}\gtrsim S^{\abs{\v n}}$.
%
% The sudden and drastic departure from virtually exact results occurs
% because the TST expansion neglects high-weight
% (i.e.~large-$\abs{\v m}$) operators $\hat\S_{\v m}$ whose
% contributions to a Heisenberg operator $\hat\O\p{t}$ of interest
% eventually become non-negligible.  For sufficiently large times $t$,
% these individual contributions can become comparable in magnitude to
% the expectation value $\bk{\hat\O\p{t}}$, and will grow with some
% large power of $t$, so their truncation will have a drastic effect
% on the estimated value of $\bk{\hat\O\p{t}}$.  This breakdown
% mechanism is also the reason for prolonged agreement between the TST
% expansion and benchmarking methods when squeezing the initial state
% $\ket{-\Z}$ rather than $\ket{\X}$: for initial states $\ket{-\Z}$,
% all initial-time correlators $\bk{\hat\S_{\v m}}$ vanish unless
% $m_+=m_-=0$ (see Appendix \ref{sec:initial_conditions}), so there is
% a substantially smaller number of neglected terms with non-zero
% contribution to the correlator $\bk{\hat\O\p{t}}$.
%
Going up through order $M=70$ in the TST expansion does not
significantly increase the breakdown time $t_{\t{break}}^{(M)}$, and
in some cases even shortens $t_{\t{break}}^{(M)}$.  The sudden and
drastic departure from virtually exact results is consistent with the
limitations of the TST expansion discussed at the end of Section
\ref{sec:theory}.  Specifically, we identify three possible mechanisms
for breakdown:
\begin{enumerate*}
\item a rapid growth in the order $M$ necessary for the TST expansion
  to converge,
\item the growth of numerical errors in excessively large terms of the
  TST expansion, and
\item the formal breakdown of the perturbative expansion in the time
  $t$.
\end{enumerate*}
In all of these cases, a detailed cancellation eventually ceases to
occur between large terms at high orders in the TST expansion.  These
large terms grow with the time $t$ raised to some large power (as high
as $M$), and therefore rapidly yield wildly unphysical results.  In
contrast to other approximate methods such as the cumulant
expansion\cite{meiser2010steadystate}, the TST expansion can thus
diagnose its own breakdown, which is an important feature when working
in parameter regimes that are inaccessible via other means to compute
correlators.
%

%
% Although the TST expansion breaks down at short times, it has two
% key advantages over other methods to compute collective spin
% correlators.  First, computing spin correlators with the TST
% expansion is generally much faster and requires much fewer computing
% resources than the alternatives.  The quantum trajectory Monte Carlo
% simulations performed for Figure \ref{fig:squeezing_incoherent}, for
% example, take $\sim10^4$ CPU hours to compute on standard modern
% computing hardware; the bulk of this time is spent performing sparse
% matrix-vector multiplication, leaving little room to further
% optimize serial runtime.  Parallelization can reduce the actual
% runtime of these quantum trajectory simulations to $\sim10$ hours by
% running all trajectories at once, but at the cost of greatly
% increasing computing resource requirements.  The TST expansion
% results in Figure \ref{fig:squeezing_incoherent}, meanwhile, take
% $\sim10$ seconds to compute with a single CPU on similar hardware.
%
Although the TST expansion breaks down at short times, it has two key
advantages over the quantum trajectory Monte Carlo method to compute
correlators in the presence of decoherence.  First, computing spin
correlators with the TST expansion is generally faster and requires
fewer computing resources.  The TST expansion results in Figure
\ref{fig:squeezing_incoherent}, for example, take $\sim10$ seconds to
compute with a single CPU on modern computing hardware.  The quantum
trajectory Monte Carlo results in the same figure, meanwhile, take
$\sim10^4$ CPU hours to compute on similar hardware; the bulk of this
time is spent performing sparse matrix-vector multiplication, leaving
little room to further optimize serial runtime.  Parallelization can
reduce actual runtime of the Monte Carlo simulations to $\sim10$ hours
by running all trajectories at once, but at the cost of greatly
increasing computing resource requirements.  Though it may be possible
to further speed up quantum trajectory Monte Carlo simulations by
introducing new truncation schemes, any modifications
\begin{enumerate*}
\item should be made carefully to ensure that simulations still yield
  correct results, and
\item are unlikely to bridge the orders of magnitude in computing
  resource requirements.
\end{enumerate*}
%

@@

\section{Conclusions}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
\label{sec:theory}

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider a system of $N$ distinct
spin-1/2 particles.  Defining individual spin-1/2 operators
$\hat s_{\a=\x,\y,\z}\equiv\hat\sigma_\a/2$ and
$\hat s_\pm\equiv\hat s_\x\pm i\hat s_\y=\hat\sigma_\pm$ with Pauli
operators $\hat\sigma_\a$, we denote an operator that acts with
$\hat s_\a$ on the spin indexed by $j$ and trivially (i.e.~with the
identity $\hat\1$) on all other spins by $\hat s_\a^{(j)}$.  We then
define the collective spin operators
$\hat S_\a\equiv\sum_{j=1}^N\hat s_\a^{(j)}$ for
$\a\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\hat\S_{\v m}}$
as a basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\hat\S_{\v m}\equiv \hat S_+^{m_+} \hat S_\z^{m_\z} \hat S_-^{m_-}$,
we can expand any collective spin operator $\hat\O$ in the form
\begin{align}
  \hat\O = \sum_{\v m} \O_{\v m} \hat\S_{\v m}
\end{align}
with scalar coefficients $\O_{\v m}\in\C$.  If $\hat\O$ is
self-adjoint, for example, then $\O_{\v m}^*=\O_{\v m^*}$ with
$\v m^*\equiv\p{m_-,m_\z,m_+}$.  The corresponding Heisenberg operator
is then
$\hat\O\p{t}=\sum_{\v m}\O_{\v m}\p{t}\hat\S_{\v m}+\hat\E_\O\p{t}$,
with time-dependent coefficients $\O_{\v m}\p{t}$ for time-independent
Schr\"odinger operators $\hat\S_{\v m}$, and mean-zero ``noise''
operators $\hat\E_\O\p{t}$ that result from interactions between the
spin system and its environment, initially $\hat\E_\O\p{0}=0$.  These
noise operators will essentially play no role in the present work, but
are necessary to include for a consistent formalism of Heisenberg
operators in an open quantum system; see Appendix \ref{sec:noise} for
further discussion.  The expectation values of Heisenberg operators
evolve according to
\begin{align}
  \f{d}{dt} \bk{\hat\O\p{t}}
  = \bk{\check T \hat\O\p{t}}
  = \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T_{\v m\v n} \O_{\v n}\p{t}
  \label{eq:EOM}
\end{align}
with a Heisenberg-picture time derivative operator $\check T=d/dt$
whose matrix elements $T_{\v m\v n}\in\C$ are defined by
\begin{align}
  \check T \hat\S_{\v n} \equiv i \sp{\hat H, \hat\S_{\v n}}_-
  + \sum_\J \gamma_\J \check\D\p{\J} \hat\S_{\v n}
  = \sum_{\v m} \hat\S_{\v m} T_{\v m\v n},
  \label{eq:time_deriv}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\hat H$ is the collective spin
Hamiltonian; $\J$ is a set of jump operators with a corresponding
decoherence rate $\gamma_\J$; and $\check\D$ is a Heisenberg-picture
dissipator, or Lindblad superoperator, defined by
\begin{align}
  \check\D\p{\J} \hat \O
  \equiv \sum_{\hat J\in\J} \p{\hat J^\dag \hat \O \hat J
    - \f12\sp{\hat J^\dag \hat J, \hat \O}_+}.
\end{align}
Decoherence via uncorrelated decay of individual spins, for example,
would be described by the set of jump operators
$\J_-\equiv\set{\hat s_-^{(j)}:j=1,2,\cdots,N}$.  The commutator in
Eq.~\eqref{eq:time_deriv} can be computed by expanding the product
$\hat\S_{\v\ell}\hat\S_{\v m} = \sum_{\v n} f_{\v\ell\v m\v
  n}\hat\S_{\v n}$ with structure constants
$f_{\v\ell\v m\v n}\in\mathbb{R}$ that we work out in Appendices
\ref{sec:identities}--\ref{sec:general_product}, and the effects of
decoherence from jump operators (i.e.~elements of $\J$) of the form
$\hat g^{(j)} = \sum_\a g_\a \hat s_\a^{(j)}$ and
$\hat G = \sum_\a G_\a \hat S_\a$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be some of the main technical
contributions of this work, with potential applications beyond the
short-time simulation method presented here.  These ingredients are
sufficient to compute matrix elements $T_{\v m\v n}$ of the time
derivative operator $\check T$ in Eq.~\eqref{eq:time_deriv} in most
cases of practical interest.

@@

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \bk{\hat\O\p{t}}
  = \bk{e^{t\check T} \hat \O\p{0}}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator $\check T^k$ are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \bk{\hat\O\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \bk{\hat\S_{\v m}} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Note that when computing an
expectation value $\bk{\hat\O\p{t}}$, the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, cuts both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}$ and the number of
matrix elements $T_{\v m\v n}$ that we may need to explicitly compute
roughly in half.

Unlike the weight-based truncation in
Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
$T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in Eq.~\eqref{eq:TST} tell us
which operators $\hat\S_{\v m}$ are relevant for computing the
expectation value $\bk{\hat\O\p{t}}$ to a fixed order $M$.  The TST
expansion thereby avoids the introduction of a weight measure $w$ that
chooses which operators to keep track of, and trades the cost of
solving a system of differential equations for the cost of computing
expectation values $\bk{\hat\S_{\v m}}$ and matrix elements
$T_{\v m\v n}^k$.  In all cases considered in this work, we find that
the TST expansion is both faster to evaluate and provides accurate
correlators $\bk{\hat\O\p{t}}$ until later times $t$ than the
weight-based expansion in \eqref{eq:weight_truncation} with weight
measure $w\p{\v m}=\abs{\v m}$ and cutoff $W\approx M$.  We therefore
restrict the remainder of our discussions to the TST expansion in
Eq.~\eqref{eq:TST}, and provide a pedagogical tutorial for computing
correlators using the TST expansion in Appendix \ref{sec:tutorial}.

Three primary considerations limit the maximum time $t$ to which we
can accurately compute a correlator $\bk{\hat S_{\v n}\p{t}}$ using
the TST expansion.  First, maintaining accuracy at larger times $t$
requires going to higher orders $M$ in the TST expansion.  An
order-$M$ TST expansion of the correlator $\bk{\hat S_{\v n}\p{t}}$
can involve a significant fraction of operators $\hat S_{\v m}$ with
weight $\abs{\v m}\lesssim M$, which implies the need to compute
$O\p{M^3}$ initial-time expectation values $\bk{\hat S_{\v m}}$ and
$O\p{M^4}$ matrix elements $T^k_{\v m\v n}$.  In practice, with a
straightforward implementation of the TST expansion we find that these
requirements generally restrict $M\lesssim 50$ -- $70$ with $8$ --
$50$ gigabytes of random access memory (RAM).  Second, individual
terms at high orders of the TST expansion in Eq.~\eqref{eq:TST} can
grow excessively large, greatly amplifying any numerical errors and
thereby spoiling cancellations that are necessary to arrive at a
physical value of a correlator, i.e.~with
$\abs{\bk{\hat S_{\v n}\p{t}}}\lesssim S^{\abs{\v n}}$.  Finally, the
TST expansion is essentially perturbative in the time $t$, which
implies that the validity of this a formal expansion eventually breaks
down.  Precisely characterizing the implications of these last two
considerations for the TST expansion requires additional analysis that
we defer to future work.  As we show from benchmarks of the TST
expansion in Section \ref{sec:squeezing}, however, a detailed
understanding of these limitations is not necessary to diagnose the
breakdown time $t_{\t{break}}^{(M)}$ beyond which the TST expansion
yields inaccurate results.  Empirically, we find that going beyond
order $M\approx35$ yields no significant gains in all cases considered
in this work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Spin squeezing, benchmarking, and breakdown}
\label{sec:squeezing}

@@

Figure \ref{fig:benchmarking} compares the squeezing parameter $\xi^2$
for $N=10^4$ spins initially in the state $\ket{\X}$ evolved under the
Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}, as computed via
both benchmarking simulations and the TST expansion in
Eq.~\eqref{eq:TST} with $M=35$.  Squeezing is shown for both unitary
dynamics (Figure \ref{fig:squeezing_unitary}), as well as non-unitary
dynamics in the presence of spontaneous decay, excitation, and
dephasing of individual spins at rates $\chi$ (Figure
\ref{fig:squeezing_incoherent}), respectively described by the sets of
jump operators $\J_\a\equiv\set{\hat s_\a^{(j)}}$ with corresponding
decoherence rates $\gamma_\a=\chi$ for $\a\in\set{-,+,\z}$.  The
results shown in Figure \ref{fig:benchmarking} were computed in a
rotated basis with
$\p{\hat s_\z,\hat s_\x}\to\p{\hat s_\x,-\hat s_\z}$ and
$\ket\X\to\ket{-\Z}\equiv\ket\dn^{\otimes N}$, as well as appropriate
transformations of the Hamiltonian and jump operators.  The only
effects of this rotation on the results presented in Figure
\ref{fig:benchmarking} are to
\begin{enumerate*}
\item reduce the time it takes to compute correlators
  $\bk{\hat\O\p{t}}$ with the TST expansion, and
\item prolong the time for which the TST expansion of TNT results
  agree with benchmarking simulations.
\end{enumerate*}
The speedup in a different basis occurs because for the initial state
$\ket{-\Z}$, all initial-time correlators $\bk{\hat S_{\v m}}$ are
zero unless $m_+=m_-=0$, and all non-zero correlators take $O\p{1}$
(i.e.~constant in $N$) time to compute, rather than $O\p{N}$ time (see
Appendix \ref{sec:initial_conditions}).  In total, the use of a
rotated basis reduces the computation time of initial-time correlators
from $O\p{M^3N}$ to $O\p{M}$.  The reason for prolonged agreement of
TNT results in a rotated basis is not entirely understood, and
provides a clue into the precise mechanism by which the TST expansion
breaks down (discussed below).  We defer a detailed study of this
breakdown to future work.

The main lesson from Figure \ref{fig:benchmarking} is that the TST
expansion yields essentially exact results right up until a sudden and
drastic departure that can be diagnosed by inspection.  The breakdown
of the TST expansion in Figure \ref{fig:benchmarking} is marked by an
unphysical squeezing parameter $\xi^2<0$, which occurs due to
individual correlators taking unphysical values with
$\abs{\bk{\hat\S_{\v n}\p{t}}}\gtrsim S^{\abs{\v n}}$.  Going up
through order $M=70$ in the TST expansion does not significantly
increase the breakdown time $t_{\t{break}}^{(M)}$, and in some cases
even shortens $t_{\t{break}}^{(M)}$.  The sudden and drastic departure
from virtually exact results is consistent with the limitations of the
TST expansion discussed at the end of Section \ref{sec:theory}.
Specifically, we identify three possible mechanisms for breakdown:
\begin{enumerate*}
\item a rapid growth in the order $M$ necessary for the TST expansion
  to converge,
\item the growth of numerical errors in excessively large terms of the
  TST expansion, and
\item the formal breakdown of the perturbative expansion in the time
  $t$.
\end{enumerate*}
In all of these cases, a detailed cancellation eventually ceases to
occur between large terms at high orders in the TST expansion.  These
large terms grow with the time $t$ raised to some large power (as high
as $M$), and therefore rapidly yield wildly unphysical results.  In
contrast to other approximate methods such as the cumulant
expansion\cite{meiser2010steadystate}, the TST expansion can thus
diagnose its own breakdown, which is an important feature when working
in parameter regimes that are inaccessible via other means to compute
correlators.

Although the TST expansion breaks down at short times, it has two key
advantages over the quantum trajectory Monte Carlo method to compute
correlators in the presence of decoherence.  First, computing spin
correlators with the TST expansion is generally faster and requires
fewer computing resources.  The TST expansion results in Figure
\ref{fig:squeezing_incoherent}, for example, take $\sim10$ seconds to
compute with a single CPU on modern computing hardware.  The quantum
trajectory Monte Carlo results in the same figure, meanwhile, take
$\sim10^4$ CPU hours to compute on similar hardware; the bulk of this
time is spent performing sparse matrix-vector multiplication, leaving
little room to further optimize serial runtime.  Parallelization can
reduce actual runtime of the Monte Carlo simulations to $\sim10$ hours
by running all trajectories at once, but at the cost of greatly
increasing computing resource requirements.  Though it may be possible
to further speed up quantum trajectory Monte Carlo simulations by
introducing new truncation schemes, any modifications
\begin{enumerate*}
\item should be made carefully to ensure that simulations still yield
  correct results, and
\item are unlikely to bridge the orders of magnitude in computing
  resource requirements.
\end{enumerate*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
